Using generalized additive models to analyze biomedical non-linear longitudinal data

### 2021-06-21 Manuscripts/Manuscript by chapters -for submission/07-Appendix.Rmd
References do not work
Need to fix the code highlighting. How to add the listing option for all chunks?

---
output:
  pdf_document: default
  html_document: default
---
# (APPENDIX) Appendix {-}

# Code for Manuscript data

This section presents the code used to generate figures, models and simulated data from Sections 3 and 4 from the main manuscript.

@@

This section simulated linear and quadratic data in the same manner as in Section \@ref(simulation). The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in  panels A and D the simulated mean responses and individual data points.  Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.  Panels D and H show the predictions from the rm-ANOVA model.

@@

```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}

##########Section for calculations###########

@@

```{r, linear-cases-Appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
# linear response, correlated errors (subject effect)
A1/B1+plot_annotation(tag_levels = 'A')
```

@@

```{r, quadratic-cases-Appendix, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
# linear response, correlated errors (subject effect)
C1/D1+plot_annotation(tag_levels = 'A')
```

@@

# Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}

@@

# GAM and Linear model plots and Missing data

@@

## GAM and Linear model plots

>>>>

# (APPENDIX) Appendix {-}

<!-- allows for page numbering using letters for the Appendix -->
\appendixpagenumbering

<!-- Allows for figure numbers with letters for Appendix -->
\counterwithin{figure}{section} 
# APPENDIX

This section presents the code used to generate figures, models and simulated data from the main manuscript.

@@

This section simulates linear and quadratic data in the same manner as in Section \@ref(simulation). The linear simulations using Figure \@ref(fig:linear-cases-Appendix) show in panels A and D the simulated mean responses and individual data points.  Panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.  Panels D and H show the predictions from the rm-ANOVA model.

@@

```{r, responses-Appendix,include=TRUE,message=FALSE,echo=TRUE}
##########Section for calculations###########

@@

```{r, linear-cases-Appendix, fig.width=14, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:linear-cases-appendix)'}
# linear response, correlated errors (subject effect)
A1/B1+plot_annotation(tag_levels = 'A')
```

@@

```{r, quadratic-cases-Appendix, fig.width=14, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap='(ref:quadratic-caption-appendix)'}
# linear response, correlated errors (subject effect)
C1/D1+plot_annotation(tag_levels = 'A')
```

@@

## Longitudinal biomedical data simulation and GAMs{#tumor-data-simulation}

@@

## GAM and Linear model plots and Missing data

@@

### GAM and Linear model plots

### 2021-06-21 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
Need to fix the code highlighting. How to add the listing option for all chunks?

---
title: Using generalized additive models to analyze biomedical non-linear longitudinal data
runninghead: Mundo \emph{et al}.
author:
- name: Ariel I. Mundo
  num: 1
- name: John R. Tipton
  num: 2
- name: Timothy J. Muldoon*
  num: 1
address:
- num: 1
  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
- num: 2
  org:  Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
corrauth: "Timothy J. Muldoon"
email: tmuldoon@uark.edu
abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research. In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
keywords: Cancer biology; Tumor response; generalized additive models; simulation; R
classoption:
  - Royal
  - times
bibliography: refs.bib
bibliographystyle: sagev
output:
  rticles::sage_article:
    keep_tex: yes
    #pandoc_args: --listings
    fig_caption: yes
    #extra_dependencies:
      #subfig: null #allows for subfigures
      #breqn: null #line breaks for long equations
      #caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
---

@@

#knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)

#global themes for plotting
thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting

#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting

#this function will remove the default "author information" that is the Pandoc LaTeX template. Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well. An explanation in detail is provided at:
#https://github.com/rstudio/bookdown/issues/1158
# remove_author <- function(x) {
#   # identify empty author line
#   i <- grep("^\\\\author\\{\\}$", x)
#   # be sure it is the one pandoc inserts
#   if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
#   x
# }
# options(bookdown.post.latex = remove_author)
```

@@

# Bibliography

<!-- <div id="refs"></div> -->

```{r child='07-Appendix.Rmd'}

```

>>>>

---
title: Using generalized additive models to analyze biomedical non-linear longitudinal data
runninghead: Mundo \emph{et al}.
author:
- name: Ariel I. Mundo
  num: 1
- name: John R. Tipton
  num: 2
- name: Timothy J. Muldoon*
  num: 1
address:
- num: 1
  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
- num: 2
  org:  Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
corrauth: "Timothy J. Muldoon"
email: tmuldoon@uark.edu
abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research. In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (two treatment groups, 10 subjects per group, five repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
keywords: Cancer biology; Tumor response; generalized additive models; simulation; R
classoption:
  - Royal
  - times
bibliography: refs.bib
bibliographystyle: sagev
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble_SAGE.sty #LaTeX formatting options
---

@@

#knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)
#global themes for plotting
thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting
```

@@

```{r child='07-Appendix.Rmd'}
```
